\section{Concluding Remarks}
\label{sec:conclusion}

We have proven that for \kl{hereditary classes} of graphs of \kl{bounded
clique-width}, being \kl{$\forall$-well-quasi-ordered} is a very robust notion,
that enjoys many structural characterisations, and where most conjectures from
the literature on labelled well-quasi-orderings hold true. We consider these
results as a first step towards understanding the landscape of labelled
well-quasi-orderings of graphs and other relational structures.

A natural research direction would be to consider other classes of relational
structures. We strongly believe that our techniques can be adapted to the case
of finite relational structures over a finite relational signature. Another
interesting direction would be to consider non-hereditary classes of graphs of
bounded clique-width. In this case, we also believe that our techniques can be
adapted.

A fundamental limitation of our work is that we assume classes to be of
\kl{bounded clique-width}. We believe that classes of graphs that are
\kl{$2$-well-quasi-ordered} are necessarily of \kl{bounded clique-width} (see
\cref{cwqo:conj}), and a first step would be to show that every
such class is \emph{monadically dependent}.

Another question left unanswered by our work is whether we can
\kl{existentially interpret} all paths in every \kl{hereditary class} of graphs
that is not \kl{$\forall$-well-quasi-ordered} (and of \kl{bounded
clique-width}). An argument similar to \cref{lem:transduction-not-wqo} shows
that if a class \kl{existentially interprets} all paths, then it is not
\kl{$2$-well-quasi-ordered}, but the converse direction remains open.

Finally, note that we proved a very strong dichotomy result for hereditary
classes of graphs of bounded clique-width: either they are not
\kl{$2$-well-quasi-ordered} or the \kl{induced subgraph} relation on them is
simpler than the \kl{gap-embedding} relation on trees. We conjecture that this
dichotomy extends to all hereditary classes of graphs: there is nothing beyond
(nested) trees.
